Draft_Application_Stochastic_Volatility

### 2021-07-21 Real Dataset SV.Rmd
update

In the following sections, we will try to replicate some of the analyses which were previously conducted on synthetic data, generated by a basic stochastic volatility model; this time, the same model specification and filters will be used on real data. However, as we will see, real data require different strategies to evaluate the effectiveness of several Sequential Monte Carlo techniques. Intuitively, such differences are due to the fact that now the latent process driving the Hidden Markov Model is not observed, nor do we know exactly its law of motion. 
This part of the analysis is structured as follows: section \textit{fill with the number of the section} provides a review of recent research in the stochastic volatility literature, including extensions of the basic specification and the use of particle filters; section \textit{fill with the number of the section} describes the data which we employ and the transformations that were applied to them. In section \textit{fill with the number of the section} we briefly compare the series of filtered states that are obtained from different particle filters; moreover, several PFs are compared in terms of computational cost in approximating the exact filtering distribution implied by our SV model. Section \textit{fill with the number of the section} features a forecast comparison between two SV specifications and a model that assumes constant mean and variance for the financial returns, namely a forecasting exercise that would not be possible were it not for particle filtering techniques. In section \textit{fill with the number of the section} we conclude by employing a non-parametric measure of realised volatility as a proxy for the latent volatility process, allowing a RMSE and MAE comparison between filters which is similar to that of section \textit{fill with the number of the section} for synthetic data.

\section{SV models + Literary Review}

@@

Note that rescaling the log returns has analogous implications for the section in which we employ measures of realised volatility as a proxy for the latent volatility process.
Let us denote by $rv5_t$ the daily realised variance at time $t$, which we obtain from the \textit{rv5} series provided in the Oxford-Man database. Now, note that in a SV framework one should not directly compare $rv5_t$ to $x_t$, as $rv5_t$ can be rather seen as an estimate of the variance of the log returns at time $t$, \textit{i.e.} of $e^{x_t}$ in the SV model.\footnote{See for example Guidolin and Pedio (2021), who employ $rv10$, another measure of realised volatility provided in the Oxford-Man Library, as a proxy for the latent conditional variance of the daily log returns on the FTSE100 index.} Moreover, by construction of $rv5_t$,\footnote{Nota sulla formula di rv5, che considera returns, non returns in percentuale.} the comparison should feature a SV model where the log returns did \textit{not} undergo the percentage transformation. Thus, in order to achieve a common scale with the percentage log returns, we need to multiply $rv5$ by $100^2$. To ease the interpretation and representation of the comparison between differently approximated filtering distributions, we finally take the square root of the rescaled $rv5$, so that we compare it to $e^{x_t/2}$.

@@

In this section the different particle filter techniques we discussed in earlier sections will be employed to approximate the filtering distribution that springs from a simple SV model applied to a real dataset. First, we provide a qualitative comparison between the filters, basing on visual analysis of the (approximated) filtering distributions. Subsequently, we move to a more quantitative approach, in order to assess the quality of the approximations as the number of particles increases. In a way, while in the first part most filters can be rather confidently assumed to be close to convergence, the second subsection will hint at possibly different computational costs across the competing filters.
Before moving to such analyses, it is worth mentioning that not all the filters we employ approximate the same filtering distribution $p(x_t|y_{1:t},\psi)$, where $\psi$ is a vector containing the parameters of the model. Indeed, the Liu and West filter actually approximates the distribution of a slightly different SV model, one that assumes an additional hierarchical level,
\begin{align*}
y_t|x_t & \sim N(0,e^{x_t})\\
x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)\\
x_0 & \sim N(0,100)\\
\alpha & \sim N(\gamma,\zeta)\\
\beta & \sim N(\pi,\phi)\\
\tau^2 & \sim IG(\nu/2,\lambda \nu/2)
\end{align*}
where $\gamma$,$\zeta$,$\pi$,$\phi$,$\nu$ and $\lambda$ are assumed to be known.\footnote{In these empirical applications we keep the values of the parameter we employed in the synthetic data part, see section \textit{fill with the number of the section}}.\

@@

In section XX we compared several filtered distributions, pointing out three main behaviours, namely that of the SIS algorithm, that of the Liu and West particle filter and that of the remaining filters. However, we did not have metrics that could help us clearly point at one behaviour as the closest to reality. In this section, we will employ a realised volatility measure as one such metric.

>>>>

In the following sections, we will try to replicate some of the analyses which were previously conducted on synthetic data, generated by a basic stochastic volatility model; this time, the same model specification and filters will be used on real data. However, as we will see, real data require different strategies to evaluate the effectiveness of several Sequential Monte Carlo techniques. Intuitively, such differences are due to the fact that now the latent process driving the Hidden Markov Model is not observed, nor do we know exactly its law of motion.\
This part of the analysis is structured as follows: section \textit{fill with the number of the section} provides a review of recent research in the stochastic volatility literature, including extensions of the basic specification and the use of particle filters; section \textit{fill with the number of the section} describes the data which we employ and the transformations that were applied to them. In section \textit{fill with the number of the section} we briefly compare the series of filtered states that are obtained from different particle filters; moreover, several PFs are compared in terms of computational cost in approximating the exact filtering distribution implied by our SV model. Section \textit{fill with the number of the section} features a forecast comparison between two SV specifications and a model that assumes constant mean and variance for the financial returns, namely a forecasting exercise that would not be possible were it not for particle filtering techniques. In section \textit{fill with the number of the section} we conclude by employing a non-parametric measure of realised volatility as a proxy for the latent volatility process, allowing a RMSE and MAE comparison between filters which is similar to that of section \textit{fill with the number of the section} for synthetic data.

\section{A review of Stochastic Volatility models}

@@

Note that rescaling the log returns has analogous implications for the section in which we employ measures of realised volatility as a proxy for the latent volatility process.
Let us denote by $rv5_t$ the daily realised variance at time $t$, which we obtain from the \textit{rv5} series provided in the Oxford-Man database. Now, note that in a SV framework one should not directly compare $rv5_t$ to $x_t$, as $rv5_t$ can be rather seen as an estimate of the variance of the log returns at time $t$, \textit{i.e.} of $e^{x_t}$ in the SV model.\footnote{See for example Guidolin and Pedio (2021), who employ $rv10$, another measure of realised volatility provided in the Oxford-Man Library, as a proxy for the latent conditional variance of the daily log returns on the FTSE100 index.} Moreover, by construction of $rv5_t$, the comparison should feature a SV model where the log returns did \textit{not} undergo the percentage transformation.\footnote{Indeed, this estimator is based on the sum of 5-minute intra-day squared returns: such returns are computed as the difference between the opening and closure price in an interval.} Thus, in order to achieve a common scale with the percentage log returns, we need to multiply $rv5$ by $100^2$. To ease the interpretation and representation of the comparison between differently approximated filtering distributions, we finally take the square root of the rescaled $rv5$, so that we compare it to $e^{x_t/2}$.

@@

In this section the different particle filter techniques we discussed in earlier sections will be employed to approximate the filtering distribution that springs from a simple SV model applied to a real dataset. First, we provide a qualitative comparison between the filters, basing on visual analysis of the (approximated) filtering distributions. Subsequently, we move to a more quantitative approach, in order to assess the quality of the approximations as the number of particles increases. In a way, while in the first part most filters can be rather confidently assumed to be close to convergence, the second subsection will hint at possibly different computational costs across the competing filters.\
Before moving to such analyses, it is worth mentioning that not all the filters we employ approximate the same filtering distribution $p(x_t|y_{1:t},\psi)$, where $\psi$ is a vector containing the parameters of the model. Indeed, the Liu and West filter actually approximates the distribution of a slightly different SV model, one that assumes an additional hierarchical level,
\begin{align*}
y_t|x_t & \sim N(0,e^{x_t})\\
x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)\\
x_0 & \sim N(0,100)\\
\alpha & \sim N(\gamma,\zeta)\\
\beta & \sim N(\pi,\phi)\\
\tau^2 & \sim IG(\nu/2,\lambda \nu/2)
\end{align*}
where $\gamma$,$\zeta$,$\pi$,$\phi$,$\nu$ and $\lambda$ are assumed to be known.\footnote{In these empirical applications we keep the values of the parameter we employed in the synthetic data part, see section \textit{fill with the number of the section}}.\

@@

In section XX we compared several filtered distributions, briefly pointing out three main behaviours, namely that of the SIS algorithm, that of the Liu and West particle filter and that of the remaining filters. However, we did not have metrics that could help us clearly point at one behaviour as the closest to reality. In this section, we will employ a realised volatility measure as one such metric.\
A rescaled realised volatility measure - the realised variance based on 5 min intraday return intervals ($rv5$ series in the Oxford-Man Realized Library) - will be used as a proxy for the latent volatility process driving the log returns on the 3 indices we presented in the previous sections.
